---
title: "DESeq2 differential expression report"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
params:
  dds: NULL
  transformed: NULL
  design: NULL
  variable: NULL
  results: NULL
  numerator: NULL
  denominator: NULL
  alpha: 0.05
  lfc: 1
  shrink: TRUE
  n_top: 50
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
library(ggplot2)
dds <- params$dds
res <- params$results
res$status <- classify_genes(res$padj, res$log2FoldChange, params$alpha, params$lfc)
pf <- S4Vectors::metadata(dds)$prefilter
colors <- c("Up" = "#C0392B", "Down" = "#2471A3",
            "Not significant" = "#B3B3B3", "Not tested (padj = NA)" = "#E0E0E0")
```

## Parameters

| Parameter | Value |
|---|---|
| Design formula | `r paste(deparse(DESeq2::design(dds)), collapse = "")` |
| Contrast | `r params$variable`: **`r params$numerator`** vs **`r params$denominator`** (reference) |
| Pre-filtering | ≥ `r pf$min_count` reads in ≥ `r pf$min_samples` samples (`r pf$n_after` / `r pf$n_before` genes kept) |
| FDR threshold | `r params$alpha` (Benjamini-Hochberg) |
| \|log2FC\| threshold | `r params$lfc` |
| LFC shrinkage | `r if (isTRUE(params$shrink)) "yes (lfcShrink, type 'normal')" else "no (MLE)"` |
| Transformation (QC) | `r S4Vectors::metadata(params$transformed)$method` |

## Samples

```{r}
knitr::kable(cbind(params$design, size_factor = round(DESeq2::sizeFactors(dds), 3)))
```

## Summary

```{r}
knitr::kable(summarise_results(res, dds))
```

## PCA

```{r}
p <- compute_pca(SummarizedExperiment::assay(params$transformed), params$design)
ggplot(p$data, aes(PC1, PC2, colour = .data[[params$variable]], label = sample)) +
  geom_point(size = 3) +
  labs(x = sprintf("PC1: %s%% variance", p$percent[1]),
       y = sprintf("PC2: %s%% variance", p$percent[2])) +
  theme_minimal()
```

## Volcano plot

```{r}
d <- res[!is.na(res$padj), ]
ggplot(d, aes(log2FoldChange, -log10(padj), colour = status)) +
  geom_point(size = 0.8, alpha = 0.7) +
  scale_colour_manual(values = colors) +
  geom_vline(xintercept = c(-params$lfc, params$lfc), linetype = "dashed") +
  geom_hline(yintercept = -log10(params$alpha), linetype = "dashed") +
  labs(x = "log2 fold change", y = "-log10 adjusted p-value", colour = NULL) +
  theme_minimal()
```

## MA plot

```{r}
ggplot(res[res$baseMean > 0, ], aes(baseMean, log2FoldChange, colour = status)) +
  geom_point(size = 0.8, alpha = 0.7) +
  scale_x_log10() +
  scale_colour_manual(values = colors) +
  geom_hline(yintercept = c(-params$lfc, params$lfc), linetype = "dashed") +
  labs(x = "Mean of normalised counts", y = "log2 fold change", colour = NULL) +
  theme_minimal()
```

## p-value distribution

```{r, fig.height = 3.5}
ggplot(res[!is.na(res$pvalue), ], aes(pvalue)) +
  geom_histogram(breaks = seq(0, 1, 0.05), fill = "#2c3e50", colour = "white") +
  labs(x = "p-value", y = "Genes") +
  theme_minimal()
```

## Top differentially expressed genes

```{r}
top <- head(res[res$status %in% c("Up", "Down"),
                c("gene", "baseMean", "log2FoldChange", "padj", "status")], 30)
top$baseMean <- round(top$baseMean, 1)
top$log2FoldChange <- round(top$log2FoldChange, 3)
top$padj <- format(top$padj, digits = 3, scientific = TRUE)
knitr::kable(top, row.names = FALSE)
```

```{r, fig.height = 8}
z <- top_gene_matrix(params$transformed, res, n = params$n_top)
if (!is.null(z)) {
  ann <- as.data.frame(params$design[colnames(z), , drop = FALSE])
  pheatmap::pheatmap(z, annotation_col = ann, show_rownames = nrow(z) <= 30,
                     color = grDevices::colorRampPalette(c("#2471A3", "white", "#C0392B"))(100),
                     main = sprintf("Top %d DE genes (row z-score)", nrow(z)))
}
```

## Session information

```{r}
sessionInfo()
```
